The relative roles of genetic, ecological and environmental factors in insular species radiations

### 2018-05-25 paper/draft.tex
comments, can input citations via betterbibtex/zotero
include some proper refs

%% major topic: besides classical mechanisms (ecological, environmental), genetic mechanisms affect speciation/radiations
Island radiations like the Darwin finches on Galápagos or \textit{Aeonium} on the Canaries have fascinated researchers throughout history %% cite? Darwin? ...?
To this date, however, it is unclear why these genera produced this multitude of species as opposed to most other genera that remained species poor.
Until recently, researchers focused on environmental and ecological factors to explain different speeds and ocurrences of speciation.
%% HERE COME SOME EXAMPLES
Yet, these mechanisms often fail to explain how the rapid establishment of reproductive barriers or isolation arises,
which ultimately results in sympatric speciation. %CLAIM! PROVIDE REFERENCES!
%% somewhere here Patino et al 2018 anagenesis, cladogenesis, etc islands
With the growing importance of genetic data and analyses, an increasing number of studies also investigates genetic processes and
their effect on population variality. %EXAMPLES
Some of these studies highlight the importance of genetic linkage or ``hitchhiking'' on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein)
as well as evolutionary rescue provided by the feedback with ecological processes (Schiffers et al. 2013).
%%TODO: here come a list of processes with examples
%% - genetic processes in general
%% - isolation processes
Besides arthropods (Howthorne \& Via 2001), these processes can be observed in plants as well. %Are you talking about islands here? Be clearer, because you said about the lack of geentic studies in island contexts.
For example, the rapid switching of pollinator syndromes in species of the genus \textit{Petunia} %use package for this!
seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits (Hermann et al.\ 2013). % cite properly!
In an island context, genetic studies are even rarer and almost exclusively of a descriptive nature.

%% Islands
Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics
like isolation (e.g. Losos \& Ricklefs 2009; Warren et al. 2015  %%Islands as model systems in ecology and evolution: prospects fifty years after MacArthur‐Wilson. EcoLett). %TODO: ref
Part of this value as study system originates by typically remarkable species.
In fact, islands often possess unique species communities and can exceede mainland areas
by a factor of 9.5 and 8.1 in plant and animal  endemism, respectively
(Kier et al. 2005. %A global assessment of endemism and species richness across island and mainland regions. PNAS). %TODO: refs
Such high endemism makes islands  especially threatened by human impacts, in particular climate change and invasive species (Fordham \& Brook 2010). %TODO: refs
Hence, understanding the origin, maintenance, functioning and evolution of island communities is of a high priority objective.
For the prediction of island community size, several theoretical frameworks have been proposed,
including one of the first and to this day still highly influential classical ecological theories,
namely MacArthur and Wilson's equilibrium theory of island biogeography (ETIB). %TODO: ref
In it, species richness on islands is the result of opposing rates of immigration and extinction,
both of which in turn are modified by island size (area) and its distance from the mainland (isolation).
This relationship was since successfully applied to a multitude of islands (Lomolina \& Weiser 2001, Triantis et al.\ 2011)%TODO: refs
, but also island like systems, such as coral reefs, lakes/ponds or forest fragments (Kelly et al.\ 1989, Peay et al.\ 2007, ). %TODO: refs
This general applicability (Lomolino 2000), however, cannot hide the fact
that the ETIB fails to formalize a mechanistic explanation of how extinction and immigration are modified by area and isolation, respectively.
Additionally, it has been demonstrated more recently that area also affects immigration via target effects (REFs) a
nd isolation affects extinction via rescue-effects (REFs).
Finally, ETIB has particularly been criticized for its lack of considering speciation (see e.g.\ Losos \& Schluter 2000). %TODO: refs
About a decade ago, island biogeography theory gained new momentum by Whittaker et al.'s General Dynamic Model (GDM, Whittaker et al.\ 2007, 2008). %TODO: ref
The conceptual framework traces island communities along the geomorphological trajectory of hotspot islands,
whith islands emerging from sea level via volcanic hotspot activity,
increasing over time in size and topographic complexity,
then eroding and subsiding in the ocean again when the island is not over the hotspot anymore.
In doing so, the model incorporates concepts of niche theory, speciation and species succession.
Yet, there are still many processes and drivers deemed important,
which are not considered by any formal theory and often little investigated
%[if you say this, you have to mention at leastr a couple os examples, otherwise it is an empty claim].

@@

>>>>

%% major topic: besides classical mechanisms (ecological, environmental), genetic mechanisms affect speciation/radiations
Island radiations like the Darwin finches on Galápagos or \textit{Aeonium} on the Canaries have fascinated researchers throughout history %% cite? Darwin? ...?
To this date, however, it is unclear why these genera produced this multitude of species as opposed to most other genera that remained species poor.
Until recently, researchers focused on environmental and ecological factors to explain different speeds and ocurrences of speciation.\cite{krugSpeciesGenusRatios2008}\cite{hurlbertProcessesGeneratingLatitudinal2014}\cite{willisSpeciesDiversityScale2002}
%% HERE COME SOME EXAMPLES
Yet, these mechanisms often fail to explain how the rapid establishment of reproductive barriers or isolation arises,
which ultimately results in sympatric speciation. %CLAIM! PROVIDE REFERENCES!
%% somewhere here Patino et al 2018 anagenesis, cladogenesis, etc islands
With the growing importance of genetic data and analyses, an increasing number of studies also investigates genetic processes and
their effect on population variability. %EXAMPLES
Some of these studies highlight the importance of genetic linkage or ``hitchhiking'' on reproductive/ecological isolation of populations \cite{hawthorneGeneticLinkageEcological2001}\cite{viaGeneticMosaicSuggests2008}%(e.g., Via \& West 2008, Via 2001 and references therein)
as well as evolutionary rescue provided by the feedback with ecological processes \cite{schiffersLimitedEvolutionaryRescue2012}.
%%TODO: here come a list of processes with examples
%% - genetic processes in general
%% - isolation processes
Besides arthropods \cite{hawthorneGeneticLinkageEcological2001}, these processes can be observed in plants as well. %Are you talking about islands here? Be clearer, because you said about the lack of geentic studies in island contexts.
For example, the rapid switching of pollinator syndromes in species of the genus \textit{Petunia} %use package for this!
seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits \cite{hermannTightGeneticLinkage2013}
In an island context, genetic studies are even rarer and almost exclusively of a descriptive nature.

%% Islands
Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics
like isolation (e.g. \cite{lososAdaptationDiversificationIslands2009b}; \cite{warrenIslandsModelSystems}.
Part of this value as study system originates by typically remarkable species.
In fact, islands often possess unique species communities and can exceede mainland areas
by a factor of 9.5 and 8.1 in plant and animal  endemism, respectively \cite{kierGlobalAssessmentEndemism2009}
Such high endemism makes islands  especially threatened by human impacts, in particular climate change and invasive species \cite{fordhamWhyTropicalIsland2010}
Hence, understanding the origin, maintenance, functioning and evolution of island communities is of a high priority objective.
For the prediction of island community size, several theoretical frameworks have been proposed,
including one of the first and to this day still highly influential classical ecological theories,
namely MacArthur and Wilson's equilibrium theory of island biogeography (ETIB). %TODO: ref
In it, species richness on islands is the result of opposing rates of immigration and extinction,
both of which in turn are modified by island size (area) and its distance from the mainland (isolation).
This relationship was since successfully applied to a multitude of islands (Lomolina \& Weiser 2001, Triantis et al.\ 2011)%TODO: refs
, but also island like systems, such as coral reefs, lakes/ponds or forest fragments (Kelly et al.\ 1989, Peay et al.\ 2007, ). %TODO: refs
This general applicability (Lomolino 2000), however, cannot hide the fact
that the ETIB fails to formalize a mechanistic explanation of how extinction and immigration are modified by area and isolation, respectively.
Additionally, it has been demonstrated more recently that area also affects immigration via target effects (REFs) a
nd isolation affects extinction via rescue-effects (REFs).
Finally, ETIB has particularly been criticized for its lack of considering speciation (see e.g.\ Losos \& Schluter 2000). %TODO: refs
About a decade ago, island biogeography theory gained new momentum by Whittaker et al.'s General Dynamic Model (GDM, Whittaker et al.\ 2007, 2008). %TODO: ref
The conceptual framework traces island communities along the geomorphological trajectory of hotspot islands,
whith islands emerging from sea level via volcanic hotspot activity,
increasing over time in size and topographic complexity,
then eroding and subsiding in the ocean again when the island is not over the hotspot anymore.
In doing so, the model incorporates concepts of niche theory, speciation and species succession.
Yet, there are still many processes and drivers deemed important,
which are not considered by any formal theory and often little investigated
%[if you say this, you have to mention at leastr a couple os examples, otherwise it is an empty claim].

@@

%Clonality
